On the presentation of orthogonal groups over polynomial rings

### 2019-06-17 dl-paper.tex
Fixed problem with symbols

\subsection{Steinberg groups}
%TODO: Describe what root systems are sympelctic and which are not
Let $\Phi$ be a reduced and irreducible root system of rank $\geq 2$ and $R$ be a commutative ring with $1$. Recall that in this case the \emph{Steinberg group} $\St(\Phi, R)$ can be defined by means of generators $x_{\alpha}(\xi)$ where $\xi\in R, \alpha\in\Phi$ and relations:
\begin{align}
& \phantom{[}
x_\alpha(s) x_\alpha(t) = x_\alpha(s+t),\ \alpha\in\Phi,\ s,t\in R; \label{rel:add}\\
& [x_\alpha(s), x_\beta(t)] = \prod\limits_{i,j\in\mathbb{N}}
 x_{i\alpha + j\beta}\left(N_{\alpha\beta ij}\, s^i t^j\right),\quad \alpha,\beta\in\Phi,\ \alpha\neq\pm\beta,\ s,t\in R. \label{rel:CCF}
\end{align}
The indices $i$, $j$ appearing in the right-hand side of the above relation range over
all positive natural numbers such that $i\alpha + j\beta\in\Phi$.
The structure constants $N_{\alpha \beta i j}=\pm 1,\pm 2,\pm 3$ appearing in \eqref{rel:CCF} depend only on $\Phi$ and {\bf TODO}:
% в литературе описываются способы выбрать эти константы ссылка на третий взгляд.

@@

In our computations we use two explicit families of elements of $\K_2(\Phi, R)$: Steinberg and Dennis--Stein symbols.
Recall that Steinberg symbols are defined for any $s, t \in R^\times$ by the identity
\begin{equation} \label{eq:steinberg} \{ s, t \}_\alpha = h_\alpha(s) \cdot h_\alpha(t) \cdot h_\alpha(st)^{-1}. \end{equation}

Our convention for Dennis--Stein symbols follows~\cite{DS73} (and {\it not} III.5.11 of ~\cite{Kbook}).
Recall that Dennis--Stein symbols are defined for arbitrary $r, s\in R$ satisfying $1 + rs \in R^\times$ as follows:
\begin{equation} \label{eq:dennis-stein}  \langle r,s \rangle _ \alpha = x_{-\alpha}\left(\tfrac{- s}{1 + rs}\right) \cdot x_{\alpha}(r) \cdot x_{-\alpha}(s) \cdot x_{\alpha}\left(\tfrac{- r}{1+rs}\right) \cdot h_{\alpha}(1 + rs)^{-1}. \end{equation}

Dennis--Stein symbols and Steinberg symbols are related to each other via the following identities:
\begin{equation} \label{DS-S-relationship} \langle a, b \rangle_\alpha = \{-a, 1+ab\}_\alpha\text{ for } a, 1+ab\in R^\times,\ \
 \{ s, t \}_\alpha = \left\langle -s, \tfrac{1 - t}{s} \right\rangle_\alpha\text{ for } s, t\in R^\times. \end{equation}

%Using~\eqref{eq:conj-h-h} it is easy to verify that \[[h_\alpha(s), h_\beta(t)] = \{s, t^{\langle \alpha, \beta \rangle}\}_\alpha = \{t, s^{\langle \beta, \alpha \rangle}\}_\beta^{-1}.\] 
It is easy to show that the above symbols depend only on the length of $\alpha$.
Moreover, if $\Phi$ is nonsymplectic, Steinberg symbols are antisymmetric and bimultiplicative, i.\,e. \begin{equation} \label{eq:symbol-properties} \{ u, st \} = \{ u, s\} \{ u, t \}, \ \{ u, v \} = \{ v, u\}^{-1}. \end{equation}

\begin{comment}
Suppose for a moment that $\langle \alpha, \beta \rangle = -1$ and  $\langle \beta, \alpha \rangle = -1$ then
\[ \{s, t^{-1} \} = \{s,  t^{-1}\}_\alpha = \{t, s^{-1} \}_\beta^{-1} = \{s^{-1}, t\} \]
In particular, $\{s, s^{-1}\} = \{s, s^{-1}\}^{-1}$ 
\end{comment}

@@

\section{First injectivity theorem}
\begin{prop} Let $\Phi$ be a reduced irreducible root system of type $\neq \rG_2$ and $k$ be arbitrary field.
Then there is a split exact sequence of abelian groups
\[ \xymatrix{ 0 \ar[r] & \K_2(\Phi, k) \ar[r] & \K_2(\Phi, k[X, X^{-1}]) \ar[r] & H(\Phi, k) \ar[r] & 0,}\text{ where} \]
\[ H(\Phi, k) = \left\{\begin{array}{ll} \K_1^\mathrm{MW}(k)& \text{if $\Phi$ is symplectic,}\\ k^* & \text{otherwise.}  \end{array}\right. \]  \end{prop}
\begin{proof} In the nonsymplectic case one can choose a long simple root $\alpha$ of $\Phi$ in such a way that there is a commutative diagram of abelian groups
\[\xymatrixcolsep{5pc}\xymatrix{k^* \ar[r]^-{\{ -, X \}_{\alpha}} \ar[rrd]_{\{-, X\}} & \K_2(\Phi, k[X, X^{-1}]) \ar[r] & \K_2(\Phi, k(X)) \ar[d]^{\cong} \\
                                                                                      &                                 & \K_2^\mathrm{M}(k(X)).} \]
Clearly, the diagonal arrow is injective by Milnor's theorem and the vertical arrow is an isomorphism by Matsumoto's theorem~\cite[Theorem~5.10]{Ma69}.
The required assertion now follow from~\cite[Satz~3]{Hur77} and the fact that the left horizontal arrow is injective.

In the symplectic case the assertion follows from~\cite[Theorem~B]{MR91}, provided one verifies that Morita--Rehmann's group $P(F)$ is isomorphic to $\K_1^\mathrm{MW}(F)$,
 which is an easy exercise. Alternatively, this isomorphism can be obtained as a corollary of~\cref{bass-ft} and~\cite[Lemma~4.1.1]{AF17}. \end{proof}

\begin{corollary} \label{field-injectivity} For $\Phi\neq\rG_2$ the map $\St(\Phi, k[X]) \to \St(\Phi, k[X, X^{-1}])$ is injective. \end{corollary}
\begin{proof} The result follows from consideration of the commutative diagram
\[\xymatrix{ & \K_2(\Phi, k) \ar[dl]_{\cong} \ar@{^{(}->}[dr] & \\
               \K_2(\Phi, k[X]) \ar[rr] &               & \K_2(\Phi, k[X, X^{-1}])} \]
in which the left arrow is an isomorphism by the corollary of~\cite[Satz~1]{Re75} and the right arrow is split injective by the above lemma. \end{proof}
\begin{theorem} \label{thm41}
Assume the notation of~\cref{lem:lemma32} and suppose additionally that $2$ is invertible in $A$.
Then for $\ell \geq 7$ the canonical map $\St(\rD_\ell, B) \to \St(\rD_\ell, R)$ is injective.
\end{theorem}
\begin{proof}
 Consider the following commutative diagram with exact rows, in which the lifting $t$ is obtained from~\cref{lem:lemma32}:
\begin{equation*} \xymatrix{
 C(\rD_\ell, B, I) \ar[r]^{j_B} \ar@{->>}[d]_{f} & \St(\rD_\ell, B, I) \ar[r]^{\mu_B} \ar[d]_{g} &
 \St(\rD_\ell, B) \ar[r]^{\pi_B} \ar[d]_{h} & \St(\rD_\ell, k[X]) \ar[d]_{i} \\
 C(\rD_\ell, R, I) \ar[r]^{j_R}         & \St(\rD_\ell, R, I) \ar[r]^{\mu_R} \ar@{-->}[ur]^{t}&
 \St(\rD_\ell, R) \ar[r]^-{\pi_R}        & \St(\rD_\ell, k[X, X^{-1}]).
}\end{equation*}
Let $a$ be an element of $\Ker(h)$. Since $i$ is injective by~\cref{field-injectivity}, the element $a$
 also lies in $\Ker(\pi_B)$ and hence comes from some $b \in \St(\rD_\ell, B, I)$ via $\mu_B$.
Since $g(b) \in \Ker(\mu_R)$ there exists some $c \in C(\rD_\ell, R, I)$ such that $j_R(c) = g(b)$. 
By~\cref{lem:prop41}, $f$ is surjective, therefore $c = f(d)$ for some $d \in C(\rD_\ell, R, I)$ and we obtain the required assertion:
 \[ 1 = \mu_Bj_B(d) = tgj_B(d) = tj_Rf(d) =t(g(b)) = \mu_B(b) = a. \qedhere \]
\end{proof}

@@

\begin{df} Define the map $S_\alpha(a, -) \colon P_\alpha(*) \to \St(\Phi, A[X, X^{-1}])$ on each coset $P_\alpha(m)$, $m \in M$ via the following formula:
\[ S_\alpha(a, g) = x_\alpha(aX^{-1})\cdot g \cdot x_\alpha\left(-\tfrac{aX^{-1}}{1 + am}\right) \cdot \{X, 1+ am\}_\alpha.\] \end{df}
It follows immediately from the definition of $S_\alpha(a, -)$ that for $g_1 \in P_\alpha(m)$, $g_2 \in P_\alpha(*)$ holds
\begin{equation} \label{eq:Smult} S_\alpha(a, g_1\cdot g_2) = S_\alpha(a, g_1) \cdot S_\alpha\left(\tfrac{a}{1+am}, g_2\right).\end{equation}

@@

\begin{lemma} \label{lem:Tulenbayev-formula} 
One has $S_\alpha(a, x_{-\alpha}(mX)) = x_{-\alpha}\left(\tfrac{mX}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot h_\alpha(1+am).$
\end{lemma}
\begin{proof} Since $\Phi$ is nonsymplectic, we can choose $\gamma \in \Phi$ such that $\langle \alpha, \gamma \rangle = -1$. Direct computation using~\eqref{eq:conj-h-x},\eqref{eq:conj-h-h},\eqref{eq:steinberg},\eqref{eq:dennis-stein} shows that
 \begin{multline*}
 x_\alpha(aX^{-1}) \cdot x_{-\alpha}(mX) = {}^{h_\gamma(X)}(x_\alpha(a) \cdot x_{-\alpha}(m)) = \\
 = {}^{h_\gamma(X)}\left( x_{-\alpha}\left(\tfrac{m}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot h_\alpha(1+am) \cdot x_\alpha\left(\tfrac{a}{1+am}\right) \right) = \\
 = x_{-\alpha}\left(\tfrac{mX}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot h_\alpha(X^{-1}(1+am))\cdot h_\alpha(X^{-1})^{-1} \cdot x_{\alpha}\left(\tfrac{aX^{-1}}{1+am}\right) = \\
 = x_{-\alpha}\left(\tfrac{mX}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot \{1+am, X^{-1}\}^{-1} \cdot h_\alpha(1+am)\cdot x_{\alpha}\left(\tfrac{aX^{-1}}{1+am}\right) = \\
 = x_{-\alpha}\left(\tfrac{mX}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot h_\alpha(1+am) \cdot x_{\alpha}\left(\tfrac{aX^{-1}}{1+am}\right) \cdot \{1+am, X\}, \end{multline*}
which implies the assertion. \end{proof}

@@

\end{comment}
\include{rehmann}

>>>>

\subsection{Steinberg groups}
Let $\Phi$ be a reduced and irreducible root system of rank $\geq 2$ and $R$ be a commutative ring with $1$. Recall that in this case the \emph{Steinberg group} $\St(\Phi, R)$ can be defined by means of generators $x_{\alpha}(\xi)$ where $\xi\in R, \alpha\in\Phi$ and relations:
\begin{align}
& \phantom{[}
x_\alpha(s) x_\alpha(t) = x_\alpha(s+t),\ \alpha\in\Phi,\ s,t\in R; \label{rel:add}\\
& [x_\alpha(s), x_\beta(t)] = \prod\limits_{i,j\in\mathbb{N}}
 x_{i\alpha + j\beta}\left(N_{\alpha\beta ij}\, s^i t^j\right),\quad \alpha,\beta\in\Phi,\ \alpha\neq\pm\beta,\ s,t\in R. \label{rel:CCF}
\end{align}
The indices $i$, $j$ appearing in the right-hand side of the above relation range over
all positive natural numbers such that $i\alpha + j\beta\in\Phi$.
The structure constants $N_{\alpha \beta i j}=\pm 1,\pm 2,\pm 3$ appearing in \eqref{rel:CCF} depend only on $\Phi$ and {\bf TODO}:
% в литературе описываются способы выбрать эти константы ссылка на третий взгляд.

@@

\subsection{$\K_2$-groups and symbols}

In our computations we use two families of explicit elements of $\K_2(\Phi, R)$ called {\it Steinberg and Dennis--Stein symbols}. Notice that our definition of these symbols follows~\cite{DS73}, rather than more modern textbooks such as~\cite{Kbook}. Recall that Steinberg symbols are defined for any $s, t \in R^\times$ by the identity
\begin{equation} \label{eq:steinberg} \{ s, t \}_\alpha = h_\alpha(st) \cdot h_\alpha^{-1}(s) \cdot h_\alpha^{-1}(t). \end{equation}
In turn, Dennis--Stein symbols are defined for arbitrary $a, b\in R$ satisfying $1 + ab \in R^\times$ by the following formula:
\begin{equation} \label{eq:dennis-stein}  \langle a,b \rangle _ \alpha = x_{-\alpha}\left(\tfrac{- b}{1 + ab}\right) \cdot x_{\alpha}(a) \cdot x_{-\alpha}(b) \cdot x_{\alpha}\left(\tfrac{- a}{1+ab}\right) \cdot h_{\alpha}^{-1}(1 + ab). \end{equation} 
Dennis--Stein symbol $\langle a, b \rangle_\alpha$ can be expressed through Steinberg symbols in the special case when either $a$ or $b$ is an invertible element of $R$. More specifically, the following formulae hold (cf.~\cite[p.~250]{DS73}).
\begin{equation} \label{DS-S-relationship} \langle a, b \rangle_\alpha = \{-a, 1+ab\}_\alpha\text{ for } a, 1+ab\in R^\times,\ \
 \{ s, t \}_\alpha = \left\langle -s, \tfrac{1 - t}{s} \right\rangle_\alpha\text{ for } s, t\in R^\times. \end{equation}
%Using~\eqref{eq:conj-h-h} it is easy to verify that \[[h_\alpha(s), h_\beta(t)] = \{s, t^{\langle \alpha, \beta \rangle}\}_\alpha = \{t, s^{\langle \beta, \alpha \rangle}\}_\beta^{-1}.\] 
Steinberg and Dennis--Stein symbols depend only on the length of $\alpha$, in particular they do not depend on $\alpha$ if $\Phi$ is simply-laced. If $\Phi$ happens to be {\it nonsymplectic}, i.\,e. $\Phi \neq \rA_1, \rB_2, \rC_{\geq 3}$, Steinberg symbols are antisymmetric and bimultiplicative, i.\,e. they satisfy the following identities: \begin{equation} \label{eq:symbol-properties} \{ u, st \} = \{ u, s\} \{ u, t \}, \ \{ u, v \} = \{ v, u\}^{-1}. \end{equation}
For these and other properties of symbols we refer the reader to~\cite{DS73}.

\begin{comment}
Suppose for a moment that $\langle \alpha, \beta \rangle = -1$ and  $\langle \beta, \alpha \rangle = -1$ then
\[ \{s, t^{-1} \} = \{s,  t^{-1}\}_\alpha = \{t, s^{-1} \}_\beta^{-1} = \{s^{-1}, t\} \]
In particular, $\{s, s^{-1}\} = \{s, s^{-1}\}^{-1}$ 
\end{comment}
The classical Matsumoto theorem (see~\cite[Theorem~5.10]{Ma69}) allows one to compute the group $\K_2(\Phi, R)$ in the special case when $R=k$ is a field. More concretely, it asserts that $\K_2(\Phi, k) \cong \K_2^\mathrm{MW}(k)$ provided $\Phi$ is symplectic and  $\K_2(\Phi, k) \cong \K_2^\mathrm{M}(k)$ otherwise. In the following lemma we compute the group $\K_2(\Phi, R)$ in yet another case when $R=k[X, X\inv]$.
\begin{lemma}\label{K2-laurent-field} Let $\Phi$ be a reduced irreducible root system of type $\neq \rG_2$ and $k$ be arbitrary field. Then there is a split exact sequence of abelian groups
\[ \xymatrix{ 0 \ar[r] & \K_2(\Phi, k) \ar[r] & \K_2(\Phi, k[X, X^{-1}]) \ar[r] & H(\Phi, k) \ar[r] & 0,}\text{ in which} \]
\[ H(\Phi, k) = \left\{\begin{array}{ll} \K_1^\mathrm{MW}(k)& \text{if $\Phi$ is symplectic,}\\ k^* & \text{otherwise.}  \end{array}\right. \]  \end{lemma}
\begin{proof} If $\Phi$ is nonsymplectic one can choose a long simple root $\alpha\in \Phi$ in such a way that there is a commutative diagram of abelian groups
\[\xymatrixcolsep{5pc}\xymatrix{k^* \ar[r]^-{\{ -, X \}_{\alpha}} \ar[rrd]_{\{-, X\}} & \K_2(\Phi, k[X, X^{-1}]) \ar[r] & \K_2(\Phi, k(X)) \ar[d]^{\cong} \\
                                                                                      &                                 & \K_2^\mathrm{M}(k(X)).} \]
Clearly, the diagonal arrow is injective by Milnor's theorem and the vertical arrow is an isomorphism by Matsumoto theorem. The assertion now follow from~\cite[Satz~3]{Hur77} and the fact that the left horizontal arrow is injective.

If $\Phi$ is symplectic the assertion follows from~\cite[Theorem~B]{MR91}, provided one verifies that Morita--Rehmann's group $P(F)$ is isomorphic to $\K_1^\mathrm{MW}(F)$,
 which is an easy exercise. Alternatively, this isomorphism can be obtained as a corollary of~\cref{bass-ft} and~\cite[Lemma~4.1.1]{AF17}. \end{proof}

@@

\section{First injectivity theorem}
\begin{corollary} \label{field-injectivity} For $\Phi\neq\rG_2$ the map $\St(\Phi, k[X]) \to \St(\Phi, k[X, X^{-1}])$ is injective. \end{corollary}
\begin{proof} The result follows from consideration of the commutative diagram
\[\xymatrix{ & \K_2(\Phi, k) \ar[dl]_{\cong} \ar@{^{(}->}[dr] & \\
               \K_2(\Phi, k[X]) \ar[rr] &               & \K_2(\Phi, k[X, X^{-1}])} \]
in which the left arrow is an isomorphism by the corollary of~\cite[Satz~1]{Re75} and the right arrow is split injective by~\cref{K2-laurent-field}. \end{proof}
\begin{theorem} \label{thm41}
Assume the notation of~\cref{lem:lemma32} and suppose additionally that $2$ is invertible in $A$.
Then for $\ell \geq 7$ the canonical map $\St(\rD_\ell, B) \to \St(\rD_\ell, R)$ is injective.
\end{theorem}
\begin{proof}
 Consider the following commutative diagram with exact rows, in which the lifting $t$ is obtained from~\cref{lem:lemma32}:
\begin{equation*} \xymatrix{
 C(\rD_\ell, B, I) \ar[r]^{j_B} \ar@{->>}[d]_{f} & \St(\rD_\ell, B, I) \ar[r]^{\mu_B} \ar[d]_{g} &
 \St(\rD_\ell, B) \ar[r]^{\pi_B} \ar[d]_{h} & \St(\rD_\ell, k[X]) \ar[d]_{i} \\
 C(\rD_\ell, R, I) \ar[r]^{j_R}         & \St(\rD_\ell, R, I) \ar[r]^{\mu_R} \ar@{-->}[ur]^{t}&
 \St(\rD_\ell, R) \ar[r]^-{\pi_R}        & \St(\rD_\ell, k[X, X^{-1}]).
}\end{equation*}
Let $a$ be an element of $\Ker(h)$. Since $i$ is injective by~\cref{field-injectivity}, the element $a$
 also lies in $\Ker(\pi_B)$ and hence comes from some $b \in \St(\rD_\ell, B, I)$ via $\mu_B$.
Since $g(b) \in \Ker(\mu_R)$ there exists some $c \in C(\rD_\ell, R, I)$ such that $j_R(c) = g(b)$. 
By~\cref{lem:prop41}, $f$ is surjective, therefore $c = f(d)$ for some $d \in C(\rD_\ell, R, I)$ and we obtain the required assertion:
 \[ 1 = \mu_Bj_B(d) = tgj_B(d) = tj_Rf(d) =t(g(b)) = \mu_B(b) = a. \qedhere \]
\end{proof}

@@

\begin{df} Define the map $S_\alpha(a, -) \colon P_\alpha(*) \to \St(\Phi, A[X, X^{-1}])$ on each coset $P_\alpha(m)$, $m \in M$ via the following formula:
\[ S_\alpha(a, g) = x_\alpha(aX^{-1})\cdot g \cdot x_\alpha\left(-\tfrac{aX^{-1}}{1 + am}\right) \cdot \{X, 1+ am\}.\] \end{df}
It follows immediately from the definition of $S_\alpha(a, -)$ that for $g_1 \in P_\alpha(m)$, $g_2 \in P_\alpha(*)$ holds
\begin{equation} \label{eq:Smult} S_\alpha(a, g_1\cdot g_2) = S_\alpha(a, g_1) \cdot S_\alpha\left(\tfrac{a}{1+am}, g_2\right).\end{equation}

@@

\begin{lemma} \label{lem:Tulenbayev-formula} 
One has $S_\alpha(a, x_{-\alpha}(mX)) = x_{-\alpha}\left(\tfrac{mX}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot h_\alpha(1+am).$
\end{lemma}
\begin{proof} Since $\Phi$ is nonsymplectic, we can choose $\gamma \in \Phi$ such that $\langle \alpha, \gamma \rangle = -1$. Direct computation using~\eqref{eq:conj-h-x},\eqref{eq:conj-h-h},\eqref{eq:steinberg},\eqref{eq:dennis-stein},\eqref{eq:symbol-properties} shows that
 \begin{multline*}
 x_\alpha(aX^{-1}) \cdot x_{-\alpha}(mX) = {}^{h_\gamma(X)}(x_\alpha(a) \cdot x_{-\alpha}(m)) = \\
 = {}^{h_\gamma(X)}\left( x_{-\alpha}\left(\tfrac{m}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot h_\alpha(1+am) \cdot x_\alpha\left(\tfrac{a}{1+am}\right) \right) = \\
 = x_{-\alpha}\left(\tfrac{mX}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot h_\alpha(X^{-1}(1+am))\cdot h_\alpha^{-1}\left(X^{-1}\right) \cdot x_{\alpha}\left(\tfrac{aX^{-1}}{1+am}\right) = \\
 = x_{-\alpha}\left(\tfrac{mX}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot \{X^{-1}, 1+am\} \cdot h_\alpha(1+am)\cdot x_{\alpha}\left(\tfrac{aX^{-1}}{1+am}\right) = \\
 = x_{-\alpha}\left(\tfrac{mX}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot h_\alpha(1+am) \cdot x_{\alpha}\left(\tfrac{aX^{-1}}{1+am}\right) \cdot \{1+am, X\}, \end{multline*}
which implies the assertion. \end{proof}

@@

\include{rehmann}
\end{comment}
